Bees need larger brains to thrive in urban environments

### 2023-04-17 Manuscript/Ms.Rmd
Update on ms from google doc

The rapid conversion of natural habitats to anthropogenic landscapes is threatening insect pollinators worldwide, raising concern on the consequences for their fundamental role as plant pollinators in natural and agricultural systems. However, not all pollinators are negatively affected by habitat conversion, as some find in anthropogenic landscapes appropriate resources to survive and reproduce. Why some animals can thrive in anthropogenic environments that most species cannot tolerate remains insufficiently understood, but it might in part reflect enhanced behavioural plasticity to exploit new resource opportunities. This theory -known as the cognitive buffer hypothesis- has received ample support in studies of vertebrates. However, it is less obvious whether the tiny brain of insects can similarly buffer individuals against the many challenges of anthropogenic habitats. To address this issue, we measured brains for `r sum(d$individuals)` individuals from `r nlevels(factor(unique(d$Species)))` species of bees, the most important group of pollinators, and evaluated the association of their brain size with detailed information on habitat preference. Our analyses revealed that bees that prefer forested or agricultural habitats had smaller brains relative to their body size, but those who prefer urban habitats tended to have larger relative brain size. In addition, urban bees also tend to have larger body sizes, and hence, show larger absolute brain sizes. Our results provide the first empirical support of the cognitive buffer hypothesis in invertebrates, suggesting that a large brain in bees could confer behavioural advantages to tolerate urban environments.

Keywords: relative brain size, habitat preference, Apoidea, urbanization.

@@

Human-dominated habitats ---notably cities--- drastically modify the original conditions where pollinators evolved, but can also offer unique ecological opportunities in the form of new nesting spots, reduced predation pressure and high food availability associated with non-indigenous plants [@cane2006; @winfree2011; @collado2019]. The question is then why only some species are able to tolerate urban environments. The “cognitive buffer hypothesis” provides an explanation for this conundrum, suggesting that in novel environments the chances to successfully survive and reproduce depends on cognition [@sol2009], that is, the processes involved in gathering, storing and reacting to environmental information [@shettleworth2009]. While the “cognitive buffer hypothesis” receives ample support from studies of vertebrates [@sol2008; @amiel2011; @sayol2016; @howell2023], similar evidence is lacking for insects.

Insects have been long associated with the most basic types of learning due to their miniature brain [@giurfa2015; @perry2017]. However, numerous studies have shown that bees and other insects have sophisticated cognition [@chittka2009; @avargues2011; @sheehan2011]. For instance, some social insects can learn by observing other individuals how to use tools to solve complex tasks [@alem2016]. In addition, there is also evidence for substantial variation across species in brain size—both in absolute terms and relative to body size— with species that have larger brains also exhibiting enhanced cognitive performance [@collado2021]. Therefore, we can ask whether the varying success of insects in human-altered habitats might be explained by variation in brain size.

@@

Our dataset contains measurements of brain and body size for bee specimens captured on flowers by hand netting in different areas of the East Coast of the United States and Europe (Spain and the Netherlands). The dataset includes information of `r sum(d$individuals)` female individuals from `r nlevels(factor(unique(d$Species)))` species that represent `r nlevels(factor(unique(tax$family)))` families and `r nlevels(factor(unique(tax$genus)))` genera. We only considered female specimens because: (i) females are involved in a greater diversity of tasks than males, and hence are expected to experience greater selective pressures from the environment; and (ii), there are evident behavioural and morphological differences between sexes that are associated with the size and structure of the brain [@streinzer2013; @roselino2015]. Brain size was measured as the weight of fixed brains and body size as intertegular span (see supplementary text). Because brain size scales allometrically with body size, we considered for our analysis the size of the brain relative to the body.

@@

Habitat preference varied substantially across species (Figure 1B) but showed none (i.e., preference for agricultural and rural habitats) or moderate phylogenetic signal (i.e., preference for natural habitat; $\lambda$ = `r lambda_nat`; *P* = `r p_nat`). In general, most bee species preferred one or two habitat types but more rarely occurred indistinctly in the three habitats (Figure 1B). The most preferred habitats were the agricultural and natural ones with `r pref_agr` and `r pref_nat` species over 80% of the values from null models, respectively (Figure 1B and Figure S3). In contrast, most species avoided or had low preference for urban habitats (`r pref_urb1` species under the 20th percentile) and just `r pref_urb2` species showed high preference for this habitat type (Figure 1B and Figure S3). The resulting distribution of habitat preferences for each of the habitats analysed followed a zero-one inflated beta distribution (Figure S3) but note that for our analyses we only modelled the extremes of the distribution (i.e., species with statistical avoidance or preference). Hence, there were high frequencies of habitat preferences close to 0 or 1 but low frequencies of intermediate values between 0 and 1.

@@

Although our model evaluating how relative brain size was associated with habitat preference explained little of the overall variance (Bayesian R$^2$ = `r r2_main`), we found evident differences when we explored this association by habitat type (Figure 2A). Specifically, we found that bees with larger brains relative to body size showed a higher preference for urban habitats than bees with smaller relative brain sizes (Figure 2A). Contrarily, bees with smaller relative brains showed higher preference in natural and agricultural habitats than bees with larger relative brain sizes (Figure 2A). Although the models of the association between brain size and intertegular span captured also little of the overall variance (Bayesian R$^2$ = `r r2_brain`; Bayesian R$^2$ = `r r2_its`, respectively), we found more marked differences in this case between habitat types (Figures 2B and 2C). Specifically, we found that bees with larger brains and body sizes tend to appear more in urban habitats and bees with smaller brains and body sizes are more common in natural and agricultural ones (Figures 2B and 2C). These findings were consistent with the analogous analyses separated by geographical regions (United States and Europe; Figure S4).

@@

The exact mechanism that allows large-brained bee species to prefer urban habitats is still unclear. However,  there are some examples of novel behaviours that might be needed to survive in urban environments, such as the use of human-made materials to nest [@allasino2019] or being able to collect pollen and nectar from exotic flowers [@lowenstein2019]. Despite the lack of broad consensus of the functional consequences of urbanisation [@buchholz2020], there is support to the idea that cities select against small body sizes [@ahrne2009; @theodorou2021]. We found support for this functional consequence with occurrence data from two different continents and over 100 bee species. This may be a consequence of the patchy resource distribution as the foraging distance in bees is positively associated with body size [@greenleaf2007; @kendall2019]. Further, the patchiness of urban resources might also favour species that are already adapted to look for target resources (i.e. specialist bees), which tend to have relatively larger brains [@sayol2020]. However, this hypothesis would need to be properly tested for bees, as there are shreds of evidence that associate increasing urbanisation with wider ecological niches [@geslin2013; @buchholz2020; @prendergast2022].

Our analyses focused on brain size mainly for reasons of data availability. The use of brain size is based on the existence of previous evidence that enlarged brains enhance cognitive performance in bees [@collado2021] and is justified because the brain is less subject
to measurement error or context-dependent biases than are experimental measures of cognition [@kamil1987]. However, the use of brain size as a proxy for cognitive performance is not exempt from criticism [@chittka2009; @lihoreau2012]. Thus, finer brain measurements like neuropils size or mushroom bodies [the suspected centres of cognitive processes in bees\; @barth1997; @withers1995; @fahrbach1995] are likely to help improve our understanding of bee neurobiology. Overall, our findings highlight the importance of the behavioural responses for understanding the dynamics of insect populations in altered environments, and stresses the need to avoid viewing them as passive agents of external factors.

@@

# Funding

@@

>>>>

The rapid conversion of natural habitats to anthropogenic landscapes is threatening insect pollinators worldwide, raising concern on the consequences for their fundamental role as plant pollinators in natural and agricultural systems. However, not all pollinators are negatively affected by habitat conversion, as some find in anthropogenic landscapes appropriate resources to survive and reproduce. Why some animals can thrive in anthropogenic environments that most species cannot tolerate remains insufficiently understood, but it might in part reflect enhanced behavioural plasticity to exploit new resource opportunities. This theory -known as the cognitive buffer hypothesis- has received ample support in studies of vertebrates. However, it is less obvious whether the tiny brain of insects can similarly buffer individuals against the many challenges of anthropogenic habitats. To address this issue, we measured brains for `r sum(d$individuals)` individuals from `r nlevels(factor(unique(d$Species)))` species of bees, the most important group of pollinators, and evaluated the association of their brain sizes with detailed information on habitat preferences. Our analyses revealed that bees that prefer forested or agricultural habitats had smaller brains relative to their body size than those who prefer urban habitats. In addition, urban bees also tend to have larger body sizes, and hence, show larger absolute brain sizes. Our results provide the first empirical support of the cognitive buffer hypothesis in invertebrates, suggesting that a large brain in bees could confer behavioural advantages to tolerate urban environments.

Keywords: relative brain size, habitat preference, Apoidea, urbanisation.
[a]

@@

Human-dominated habitats ---notably cities--- drastically modify the original conditions where pollinators evolved, but can also offer unique ecological opportunities in the form of new nesting spots, reduced predation pressure and high food availability associated with non-indigenous plants [@cane2006; @winfree2011; @collado2019]. The question is then why only some species are able to tolerate urban environments. The “cognitive buffer hypothesis” provides an explanation for this conundrum, suggesting that in novel environments the chances to successfully survive and reproduce depends on cognition [@sol2009], that is, the processes involved in gathering, storing and reacting to environmental information [@shettleworth2009]. While the “cognitive buffer hypothesis” receives ample support from studies of vertebrates [@sol2008; @amiel2011; @sayol2016; @howell2023], similar evidence is lacking for insects. 
[b]

Insects have been long associated with the most basic types of learning due to their miniature brain [@giurfa2015; @perry2017]. However, numerous studies have shown that bees and other insects have sophisticated cognition [@chittka2009; @avargues2011; @sheehan2011]. For instance, some social insects can learn by observing other individuals how to use tools to solve complex tasks [@alem2016][c]. In addition, there is also evidence for substantial variation across species in brain size—both in absolute terms and relative to body size— with species that have larger brains also exhibiting enhanced cognitive performance [@collado2021]. Therefore, we can ask whether the varying success of insects in human-altered habitats might be explained by variation in brain size.

@@

Our dataset contains measurements of brain and body size for bee specimens captured on flowers by hand netting in different areas of the East Coast of the United States and Europe (Spain and the Netherlands). The dataset includes information of `r sum(d$individuals)` female individuals from `r nlevels(factor(unique(d$Species)))` species that represent `r nlevels(factor(unique(tax$family)))` families and `r nlevels(factor(unique(tax$genus)))` genera. We only considered female specimens because: (i) females are involved in a greater diversity of tasks than males, and hence are expected to experience greater selective pressures from the environment; and (ii), there are evident behavioural and morphological differences between sexes that are associated with the size and structure of the brain [@streinzer2013; @roselino2015]. Brain size was measured as the weight of fixed brains[d] and body size as intertegular[e] span (see supplementary text). Because brain size scales allometrically with body size, we considered for our analysis the size of the brain relative to the body.

@@

Habitat preference varied substantially across species (Figure 1B) but showed none (i.e., preference for agricultural and rural habitats) or moderate phylogenetic signal (i.e., preference for natural habitat; $\lambda$ = `r lambda_nat`; *P* = `r p_nat`). In general, most bee species preferred one or two habitat types but more rarely occurred indistinctly in the three habitats (Figure 1B). The most preferred habitats were the agricultural and natural ones with `r pref_agr` and `r pref_nat` species over 80% of the values from null models, respectively (Figure 1B and Figure S3). In contrast, most species avoided or had low preference for urban habitats (`r pref_urb1` species under the 20th percentile) and just `r pref_urb2` species showed high preference for this habitat type (Figure 1B and Figure S3). The resulting distribution of habitat preferences for each of the habitats analysed followed a zero-one inflated beta distribution (Figure S3), indicating that there were high frequencies of habitat preferences close to 0 or 1 but low frequencies of intermediate values between 0 and 1. Hence, in our analyses we only modelled the extremes of the distribution (i.e., species classified as avoiding or preferring a given habitat).

@@

Although our model evaluating how relative brain size was associated with habitat preference explained little of the overall variance (Bayesian R$^2$ = `r r2_main`), we found differences when we explored this association by habitat type (Figure 2A). Specifically, we found that bees with larger brains relative to body size showed a higher preference for urban habitats than bees with smaller relative brain sizes (Figure 2A). Contrarily, bees with smaller relative brains showed higher preference in natural and agricultural habitats than bees with larger relative brain sizes (Figure 2A). Although the models of the association between absolute brain size and intertegular span with habitat preference captured also little of the overall variance (Bayesian R$^2$ = `r r2_brain`; Bayesian R$^2$ = `r r2_its`, respectively), we found marked differences between habitat types (Figures 2B and 2C). Specifically, we found that bees with larger brains and body sizes tend to appear more in urban habitats and bees with smaller brains and body sizes are more common in natural and agricultural ones (Figures 2B and 2C). These findings were consistent with the analogous analyses separated by geographical regions (United States and Europe; Figure S4).

@@

The exact mechanism that allows large-brained bee species to prefer urban habitats is still unclear. However,  there are some examples of novel behaviours that might be needed to survive in urban environments, such as the use of human-made materials to nest [@allasino2019] or being able to collect pollen and nectar from exotic flowers [@lowenstein2019]. In addition, despite the lack of broad consensus of the functional consequences of urbanisation @buchholz2020], there is support to the idea that cities select against small body sizes [@ahrne2009; @theodorou2021]. We also found larger bees in urban areas using data from two different continents and over 100 bee species. This may be a consequence of the patchy resource distribution as the foraging distance in bees is positively associated with body size [@greenleaf2007; @kendall2019]. Further, the patchiness of urban resources might also favour species that are already adapted to look for target resources (i.e. specialist bees), which tend to have relatively larger brains [@sayol2020]. However, this hypothesis requires that the needed resources are present. In any case, this idea would need to be properly tested for bees, as there are also shreds of evidence that associate increasing urbanisation with wider ecological niches [@geslin2013; @buchholz2020; @prendergast2022].

Our analyses focused on brain size mainly for reasons of data availability. The use of brain size is based on the existence of previous evidence that enlarged brains enhance cognitive performance in bees [@collado2021] and is justified because the brain is less subject
to measurement error or context-dependent biases than are experimental measures of cognition [@kamil1987]. However, the use of brain size as a proxy for cognitive performance is not exempt from criticism [@chittka2009; @lihoreau2012]. Thus, finer brain measurements like neuropils size or mushroom bodies [the suspected centres of cognitive processes in bees\; @barth1997; @withers1995; @fahrbach1995] are likely to help improve our understanding of bee cognitive abilities in relationship to urbanization. Overall, our findings highlight the importance of behavioural responses for understanding the dynamics of insect populations in altered environments, and stresses the need to avoid viewing them as passive agents of external factors.

@@

# Funding[f][g][h]

@@

[a]Quizá completar el cupo de keywords? propongo Urban habitats, pollinators, cognitive buffer hypothesis.
[b]Tenemos alguna forma de dar pruebas de esto? Sé que demostrar la no-existencia de algo es muy difícil, pero se me queda un poco cojo afirmar esto sin nada más. No he encontrado nada. Supongo que si hay problemas con esto los revisores lo indicarán.
[c]El artículo de los bombus jugando llegó al mainstream, quizá sería buena estrategia citarlo, además así tenemos una cita de algo actual!

https://www.sciencedirect.com/science/article/pii/S0003347222002366?via%3Dihub
[d]cita sayol aqui.
[e]citar Kendall aqui
[f]Qué pondrías aquí?
[g]puedes rescatar los proyectos del mineco que agradece MA en sus papers? Survive_change, creo. Nada más por mi parte.
[h]@xmiguelangelcolladox@gmail.com podrías rescatar el funding que creas más oportuno?
